A conditional permutation-based approach to test confounder effect and center-bias in machine learning models

### 2021-10-21 manuscript.tex
Updates from Overleaf

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Validation on simulated data}

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.40\paperwidth]{fig/sim_non-norm.eps}
  \caption{\textbf{Robustness of conditional permutation based confound testing to the violation of normality assumptions for $\y$ and $\c$.} \\
  Simulations included variables with five different degrees of non-normality (A), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation, from normal (left) to severely skewed and heavy tailed (right). Fisher's kurtosis and skewness is given for each at the bottom of panel A. Panel (B) depicts false positive rates in the simulations under H0, separately for different signal-in-confounder ratios (here identical to $w_{y}$).
  Vertical axis shows type I error rate. The nominal type I error rate of $alpha=0.05$ is shown with a vertical dashed line. Panel (C) shows how violation of the normality assumption affects power. Boxplots show the mean power over simulation cases, grouped by confound-to-target ratio and sample size.}
  \label{fig:sim-non-normal}
\end{figure}

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.pdf}
  \caption{\textbf{Acquisition- and age-bias of fluid IQ prediction in the HCP dataset.} \\
  Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies. Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data. This bias is robustly detected by the proposed 'partial' confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
  Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the 'partial confound test'. P-values of the 'full' confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:hcp}
\end{figure}

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.pdf}
  \caption{\textbf{Center- and motion age-bias of ASD diagnosis prediction in the ABIDE dataset.} \\
  Boxplots and points show the predicted class probabilities (0: HC, 1: ASD), separately for the HC and ASD groups. In the top panel, predictions are plotted for each center separately and color indicates the true diagnosis (DX). AT the bottom plot, color indicates the normalized index of in-scanner motion (based on mean framewise displacement, mean FD). The proposed confounder test reveals significant center- and motion-bias in the model trained on the raw data (p<0.0001). While center bias seems to be effectively mitigated by feature regression, the same technique fails to remove motion-bias (actually introduces a paradoxical negative bias). COMBAT seems to effectively remove the effect of both confounders on the predictions and improves the predictive performance of the models in both cases.
  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the 'partial confound test'. P-values of the 'full' confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:abide}
\end{figure}

@@

% tab:unconditional-pvals
\renewcommand{\arraystretch}{1.2}
\begin{table}
\centering
\begin{tabular}{lll|ll|ll|ll|ll} 
dataset & conf. & method & $R^2_{y, c}$ & $p_{y, c}$ & $R^2_{\hat{y}, c}$ & $p_{\hat{y}, c}$ & $R^2_{\hat{y}, y}$ & $p_{\hat{y}, y}$ & partial test & full test  \\
\hline
HCP & acq.  & raw      & 0.032 & <0.001 & 0.071 & <0.001  & 0.095 & <0.001 & <0.0001 & <0.0001 \\
    &              & f.reg.    & & & 0.011 & 0.477 & 0.105  & <0.001  & 0.72 & <0.0001 \\
    &              & COMBAT    & & &0.013 & 0.4 & 0.122  & <0.001  & 0.88 & <0.0001\\
\hline
    & age   & raw       & 0.011 & 0.001  & 0.034 & <0.001  & 0.095  & <0.001 & <0.0001 & <0.0001 \\
    &       & f.reg.    & & &0.004 & 0.052 & 0.114  & <0.001  & 0.23 & <0.0001 \\
    &       & COMBAT    & & &0.005 & 0.048 & 0.121 & <0.001 & 0.17 & <0.0001 \\
\hline
ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & <0.0001 & <0.0001 \\
        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & 0.04 & <0.0001 \\
        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & 0.01 & <0.0001 \\
\hline
        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & <0.0001& <0.0001 \\
        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & <0.0001 & 0.09 \\
        &          & COMBAT    & & &  0.002 & 0.19 & 0.111     & <0.001 & 0.64 & <0.0001 \\

\end{tabular}
\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated models.}
\end{table}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/sim_h0_demo.eps}
  \caption{\textbf{Type-I error control and power of the conditional permutation-based confound testing for predictive modelling in normally distributed simulated data}. \\
  Heatmaps depict positive rates (ratio of p-values lower than 0.05, color coded as shown by the palette at the bottom) in various simulations settings (100 simulations per tile) with different covariance between the target and the confounder (given as $R^2_{y,c}$ on the vertical axes of the heatmaps), for different sample sizes (N, horizontal axes).
  Rows and columns correspond to simulations with different signal-in-prediction (y+c) and confound-to-signal values, respectively. The first row and the first column contain simulations under H0 (no confounder bias) and separated form the simulations under the alternative hypothesis with dashed line. Positive rates for the simulations under the null and alternative hypotheses can be interpreted as type I error rate and statistical power, respectively. The higher confidence limit for the type I error rates is 0.11 for each tile, with 5 tiles expected to pass it form all H0 tiles.
  The actual weights $w_{Y}$, $w_{C}$ and $w_{e}$ are depicted at the bars on top of each heatmap and color coded as shown in the legend at the bottom. Above each heatmap, the strengths of the associations of $\hat{y}$ to $Y$ and $C$, respectively, are given with $R^2$ values averaged over all simulation cases of the heatmap.
  }
  \label{fig:sim-normal}
\end{figure}

\subsection{Validation on simulated data}

@@

\begin{figure}[!b]
  \centering

  \includegraphics[width=0.40\paperwidth]{fig/sim_non-norm.eps}
  \caption{\textbf{Robustness of conditional permutation based confound testing to the violation of normality assumptions for $\y$ and $\c$.} \\
  Simulations included variables with five different degrees of non-normality (A), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation, from normal (left) to severely skewed and heavy tailed (right). Fisher's kurtosis and skewness is given for each at the bottom of panel A. Panel (B) depicts false positive rates in the simulations under H0, separately for different signal-in-confounder ratios (here identical to $w_{y}$).
  Vertical axis shows type I error rate. The nominal type I error rate of $alpha=0.05$ is shown with a vertical dashed line. Panel (C) shows how violation of the normality assumption affects power. Boxplots show the mean power over simulation cases, grouped by confound-to-target ratio and sample size.}
  \label{fig:sim-non-normal}
\end{figure}

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_hcp.png}
  \caption{\textbf{Acquisition- and age-bias of fluid IQ prediction in the HCP dataset.} \\
  Scatter plots and regression lines (with 95\% confidence intervals) show the association of the observed (horizontals axis) and predicted (vertical axis) fluid intelligence scores with various confound regression strategies. Color-coding of the confounder variables (top: acquisition batch, bottom: age group, as shown by the corresponding legends) reveals confounder-bias for both acquisition and age in the models trained on the raw data. This bias is robustly detected by the proposed 'partial' confounder test ($p<0.0001$) and seems to be effectively mitigated by both feature regression and COMBAT.
  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Both confound mitigation techniques, but especially COMBAT, improves the predictive performance of the models.
  Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the 'partial confound test'. P-values of the 'full' confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:hcp}
\end{figure}

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\paperwidth]{fig/fig_abide.png}
  \caption{\textbf{Center- and motion age-bias of ASD diagnosis prediction in the ABIDE dataset.} \\
  Boxplots and points show the predicted class probabilities (0: HC, 1: ASD), separately for the HC and ASD groups. In the top panel, predictions are plotted for each center separately and color indicates the true diagnosis (DX). AT the bottom plot, color indicates the normalized index of in-scanner motion (based on mean framewise displacement, mean FD). The proposed confounder test reveals significant center- and motion-bias in the model trained on the raw data (p<0.0001). While center bias seems to be effectively mitigated by feature regression, the same technique fails to remove motion-bias (actually introduces a paradoxical negative bias). COMBAT seems to effectively remove the effect of both confounders on the predictions and improves the predictive performance of the models in both cases.
  Relation between the observed ($\hat{y}$) and predicted IQ scores and the confounder variables is shown by the graphs as $R^2 values$. Solid red line between the confounder and the prediction means significant confounder bias, whereas blue dashed line denotes that confounder testing provided no evidence for bias. P-values are determined with the 'partial confound test'. P-values of the 'full' confound test were all less then 0.0001, i.e. confounder did not fully drive prediction for any of the models.
  }
  \label{fig:abide}
\end{figure}

@@

% tab:unconditional-pvals
\renewcommand{\arraystretch}{1.2}
\begin{table}
\centering
\begin{tabular}{lll|ll|ll|ll|ll} 
dataset & conf. & method & $R^2_{y, c}$ & $p_{y, c}$ & $R^2_{\hat{y}, c}$ & $p_{\hat{y}, c}$ & $R^2_{\hat{y}, y}$ & $p_{\hat{y}, y}$ & partial test & full test  \\
\hline
HCP & acq.  & raw      & 0.032 & <0.001 & 0.071 & <0.001  & 0.095 & <0.001 & \textbf{<0.0001} & <0.0001 \\
    &              & f.reg.    & & & 0.011 & 0.477 & 0.105  & <0.001  & 0.73 & <0.0001 \\
    &              & COMBAT    & & &0.013 & 0.4 & 0.122  & <0.001  & 0.65 & <0.0001\\
\hline
    & age   & raw       & 0.011 & 0.001  & 0.034 & <0.001  & 0.095  & <0.001 & \textbf{<0.0001} & <0.0001 \\
    &       & f.reg.    & & &0.004 & 0.052 & 0.114  & <0.001  & 0.2 & <0.0001 \\
    &       & COMBAT    & & &0.005 & 0.048 & 0.121 & <0.001 & 0.16 & <0.0001 \\
\hline
ABIDE   & center   & raw       & 0.019  & <0.001 &  0.169 & <0.001& 0.126     & <0.001 & \textbf{<0.0001} & <0.0001 \\
        &          & f.reg.    &  & &  0.042 & 0.007 & 0.173     & <0.001 & \textbf{0.04} & <0.0001 \\
        &          & COMBAT    &  & &  0.05 & 0.001 & 0.17     & <0.001 & \textbf{0.009} & <0.0001 \\
\hline
        & motion   & raw       & 0.028 & <0.001 & 0.111    &  <0.001 & 0.126    & <0.001 & \textbf{<0.0001} & <0.0001 \\
        &          & f.reg.    & & & 0.359    & <0.001  & 0.02   & <0.001 & \textbf{<0.0001} & 0.1 \\
        &          & COMBAT    & & &  0.002 & 0.19 & 0.111     & <0.001 & 0.59 & <0.0001 \\

\end{tabular}
\caption{\label{tab:unconditional-pvals} Unconditioned coefficients-of-determination, the corresponding p-values and the p-values of the partial and full confounder tests for all investigated models. Significant confoudner bias is denoted by bold font.}
\end{table}
